A Gaussian process model of human electrocorticographic data

### 2019-01-10 paper/main.tex
another reference formatting thing
updated cite to citep

An increasingly popular approach is to improve the effective spatial
resolution of MEG or scalp EEG data by using a geometric approach
called \textit{beamforming} to solve the biomagnetic or bioelectrical
inverse problem~\cite{Sarv87}.  This approach entails using detailed
brain conductance models (often informed by high spatial resolution
anatomical MRI images) along with the known sensor placements
(localized precisely in 3D space) to reconstruct brain signals
originating from theoretical point sources deep in the brain (and far
from the sensors).  Traditional beamforming approaches must overcome
two obstacles.  First, the inverse problem beamforming seeks to solve
has infinitely many solutions.  Researchers have made traction towards
constraining the solution space by assuming that signal-generating
sources are localized on a regularly spaced grid spanning the brain
and that individual sources are small relative to their distances to
the sensors~\cite{Snyd91, BailEtal01, HillEtal05}.  The second, and in
some ways much more serious, obstacle is that the magnetic fields
produced by external (noise) sources are substantially stronger than
those produced by the neuronal changes being sought (i.e., at deep
structures, as measured by sensors at the scalp).  This means that
obtaining adequate signal quality often requires averaging the
measured responses over tens to hundreds of responses or trials~\citep[e.g., see review by][]{HillEtal05}).

Another approach to obtaining high spatiotemporal resolution
neural data has been to collect fMRI and EEG data simultaneously.
Simultaneous fMRI-EEG has the potential to balance the high spatial
resolution of fMRI with the high temporal resolution of scalp EEG,
thereby, in theory, providing the best of both worlds.  In practice,
however, the signal quality of both recordings suffers substantially
when the two techniques are applied simultaneously (e.g., see review
by~\cite{HustEtal12}).  In addition, the experimental designs that are
ideally suited to each technique individually are somewhat at odds.
For example, fMRI experiments often lock stimulus presentation
events to the regularly spaced image acquisition time (TR), which
maximizes the number of post-stimulus samples.  By contrast, EEG
experiments typically employ jittered stimulus presentation times to
maximize the experimentalist's ability to distinguish electrical brain
activity from external noise sources such as from 60 Hz alternating
current power sources.

The current ``gold standard'' for precisely localizing signals and
sampling at high temporal resolution is to take (ECoG or iEEG)
recordings from implanted electrodes (but from a limited set of
locations in any given brain).  This begs the following question: what
can we infer about the activity exhibited by the rest of a person's
brain, given what we learn from the limited intracranial recordings we
have from their brain and additional recordings taken from
\textit{other} people's brains?  Here we develop an approach, which we
call \textit{SuperEEG}\footnote{The term ``SuperEEG''
  was coined by Robert J. Sawyer in his popular science fiction novel
  \textit{The Terminal Experiment}~\cite{Sawy95}.  SuperEEG is a
  fictional technology that measures ongoing neural activity
  throughout the entire living human brain with perfect precision and
  at arbitrarily high spatiotemporal resolution.}, based on Gaussian
process regression~\cite{Rasm06}.  SuperEEG entails using data from
multiple people to estimate activity patterns at arbitrary locations
in each person's brain (i.e., independent of their electrode
placements).  We test our SuperEEG approach using two large datasets
of intracranial recordings~\cite{SedeEtal03, SedeEtal07a, SedeEtal07b,
  MannEtal11, MannEtal12, EzzyEtal17, HoraEtal17, KragEtal17,
  KuceEtal17, LinEtal17, SoloEtal18, WeidEtal18, EzzyEtal18,
  KuceEtal18}.  We show that the SuperEEG algorithm recovers signals
well from electrodes that were held out of the training dataset.  We
also examine the factors that influence how accurately activity may be
estimated (recovered), which may have implications for electrode
design and placement in neurosurgical applications.

\section*{Approach}
The SuperEEG approach to inferring high temporal resolution full-brain
activity patterns is outlined and summarized in
Figure~\ref{fig:methods}. We describe (in this section) and evaluate
(in \textit{Results}) our approach using a two large previously
collected dataset comprising multi-session intracranial recordings.
Dataset 1 comprises multi-session recordings taken from 6876
electrodes implanted in the brains of 88 epilepsy
patients~\cite{SedeEtal03, SedeEtal07a, SedeEtal07b, MannEtal11,
  MannEtal12}.  Each recording session lasted from 0.2--3~h (total
recording time: 0.3--14.2~h; Fig.~\suppstats E).  During each
recording session, the patients participated in a free recall list
learning task, which lasted for up to approximately 1~h.  In addition,
the recordings included ``buffer'' time (the length varied by patient)
before and after each experimental session, during which the patients
went about their regular hospital activities (confined to their
hospital room, and primarily in bed).  These additional activities
included interactions with medical staff and family, watching
television, reading, and other similar activities.  For the purposes
of the Dataset 1 analyses presented here, we aggregated all data
across each recording session, including recordings taken during the
main experimental task as well as during non-experimental time.  We
used Dataset 1 to develop our main SuperEEG approach, and to examine
the extent to which SuperEEG might be able to generate task-general
predictions.  Dataset 2 comprised multi-session recordings from 4436
electrodes implanted in the brains of 40 epilepsy
patients~\citep{EzzyEtal17, HoraEtal17, KragEtal17, KuceEtal17,
  LinEtal17, SoloEtal18, WeidEtal18, EzzyEtal18, KuceEtal18}.  Each
recording session lasted from 0.4--2.2~h (total recording time:
0.4--6.6~h; Fig.~\suppstats K).  Whereas Dataset 1 included recordings
taken as the patients participated in a variety of activities, Dataset
2 included recordings taken as each patient performed each of two
specific experimental memory tasks: a random word list free recall
task (Experiment 1) and a categorized word list free recall task
(Experiment 2).  We used Dataset 2
to further examine the ability of SuperEEG to generalize its
predictions within versus across tasks.  Figure~\suppstats~provides
additional information about both datasets.

@@

One potential limitation of our approach is that it does not provide a
natural means of estimating the precise timing of single-neuron action
potentials.  Prior work has shown that gamma band and broadband
activity in the LFP may be used to estimate the firing rates of
neurons that underly the population contributing to the
LFP~\cite{MillEtal08, MannEtal09, JacoEtal10b, CronEtal11}.  Because
SuperEEG reconstructs LFPs throughout the brain, one could in
principle use gamma or broadband power in the reconstructed signals to
estimate the corresponding firing rates (though not the timings of
individual action potentials).

@@

\bibliography{memlab}
\bibliographystyle{CSE}

>>>>

An increasingly popular approach is to improve the effective spatial
resolution of MEG or scalp EEG data by using a geometric approach
called \textit{beamforming} to solve the biomagnetic or bioelectrical
inverse problem~\citep{Sarv87}.  This approach entails using detailed
brain conductance models (often informed by high spatial resolution
anatomical MRI images) along with the known sensor placements
(localized precisely in 3D space) to reconstruct brain signals
originating from theoretical point sources deep in the brain (and far
from the sensors).  Traditional beamforming approaches must overcome
two obstacles.  First, the inverse problem beamforming seeks to solve
has infinitely many solutions.  Researchers have made traction towards
constraining the solution space by assuming that signal-generating
sources are localized on a regularly spaced grid spanning the brain
and that individual sources are small relative to their distances to
the sensors~\citep{Snyd91, BailEtal01, HillEtal05}.  The second, and in
some ways much more serious, obstacle is that the magnetic fields
produced by external (noise) sources are substantially stronger than
those produced by the neuronal changes being sought (i.e., at deep
structures, as measured by sensors at the scalp).  This means that
obtaining adequate signal quality often requires averaging the
measured responses over tens to hundreds of responses or trials~\citep[e.g., see review by][]{HillEtal05}.

Another approach to obtaining high spatiotemporal resolution
neural data has been to collect fMRI and EEG data simultaneously.
Simultaneous fMRI-EEG has the potential to balance the high spatial
resolution of fMRI with the high temporal resolution of scalp EEG,
thereby, in theory, providing the best of both worlds.  In practice,
however, the signal quality of both recordings suffers substantially
when the two techniques are applied simultaneously~\citep[e.g., see review by][]{HustEtal12}.  In addition, the experimental designs that are
ideally suited to each technique individually are somewhat at odds.
For example, fMRI experiments often lock stimulus presentation
events to the regularly spaced image acquisition time (TR), which
maximizes the number of post-stimulus samples.  By contrast, EEG
experiments typically employ jittered stimulus presentation times to
maximize the experimentalist's ability to distinguish electrical brain
activity from external noise sources such as from 60 Hz alternating
current power sources.

The current ``gold standard'' for precisely localizing signals and
sampling at high temporal resolution is to take (ECoG or iEEG)
recordings from implanted electrodes (but from a limited set of
locations in any given brain).  This begs the following question: what
can we infer about the activity exhibited by the rest of a person's
brain, given what we learn from the limited intracranial recordings we
have from their brain and additional recordings taken from
\textit{other} people's brains?  Here we develop an approach, which we
call \textit{SuperEEG}\footnote{The term ``SuperEEG''
  was coined by Robert J. Sawyer in his popular science fiction novel
  \textit{The Terminal Experiment}~\cite{Sawy95}.  SuperEEG is a
  fictional technology that measures ongoing neural activity
  throughout the entire living human brain with perfect precision and
  at arbitrarily high spatiotemporal resolution.}, based on Gaussian
process regression~\citep{Rasm06}.  SuperEEG entails using data from
multiple people to estimate activity patterns at arbitrary locations
in each person's brain (i.e., independent of their electrode
placements).  We test our SuperEEG approach using two large datasets
of intracranial recordings~\citep{SedeEtal03, SedeEtal07a, SedeEtal07b,
  MannEtal11, MannEtal12, EzzyEtal17, HoraEtal17, KragEtal17,
  KuceEtal17, LinEtal17, SoloEtal18, WeidEtal18, EzzyEtal18,
  KuceEtal18}.  We show that the SuperEEG algorithm recovers signals
well from electrodes that were held out of the training dataset.  We
also examine the factors that influence how accurately activity may be
estimated (recovered), which may have implications for electrode
design and placement in neurosurgical applications.

\section*{Approach}
The SuperEEG approach to inferring high temporal resolution full-brain
activity patterns is outlined and summarized in
Figure~\ref{fig:methods}. We describe (in this section) and evaluate
(in \textit{Results}) our approach using a two large previously
collected dataset comprising multi-session intracranial recordings.
Dataset 1 comprises multi-session recordings taken from 6876
electrodes implanted in the brains of 88 epilepsy
patients~\citep{SedeEtal03, SedeEtal07a, SedeEtal07b, MannEtal11,
  MannEtal12}.  Each recording session lasted from 0.2--3~h (total
recording time: 0.3--14.2~h; Fig.~\suppstats E).  During each
recording session, the patients participated in a free recall list
learning task, which lasted for up to approximately 1~h.  In addition,
the recordings included ``buffer'' time (the length varied by patient)
before and after each experimental session, during which the patients
went about their regular hospital activities (confined to their
hospital room, and primarily in bed).  These additional activities
included interactions with medical staff and family, watching
television, reading, and other similar activities.  For the purposes
of the Dataset 1 analyses presented here, we aggregated all data
across each recording session, including recordings taken during the
main experimental task as well as during non-experimental time.  We
used Dataset 1 to develop our main SuperEEG approach, and to examine
the extent to which SuperEEG might be able to generate task-general
predictions.  Dataset 2 comprised multi-session recordings from 4436
electrodes implanted in the brains of 40 epilepsy
patients~\citep{EzzyEtal17, HoraEtal17, KragEtal17, KuceEtal17,
  LinEtal17, SoloEtal18, WeidEtal18, EzzyEtal18, KuceEtal18}.  Each
recording session lasted from 0.4--2.2~h (total recording time:
0.4--6.6~h; Fig.~\suppstats K).  Whereas Dataset 1 included recordings
taken as the patients participated in a variety of activities, Dataset
2 included recordings taken as each patient performed each of two
specific experimental memory tasks: a random word list free recall
task (Experiment 1) and a categorized word list free recall task
(Experiment 2).  We used Dataset 2
to further examine the ability of SuperEEG to generalize its
predictions within versus across tasks.  Figure~\suppstats~provides
additional information about both datasets.

@@

One potential limitation of our approach is that it does not provide a
natural means of estimating the precise timing of single-neuron action
potentials.  Prior work has shown that gamma band and broadband
activity in the LFP may be used to estimate the firing rates of
neurons that underly the population contributing to the
LFP~\citep{MillEtal08, MannEtal09, JacoEtal10b, CronEtal11}.  Because
SuperEEG reconstructs LFPs throughout the brain, one could in
principle use gamma or broadband power in the reconstructed signals to
estimate the corresponding firing rates (though not the timings of
individual action potentials).

@@

\bibliography{CDL-bibliography/memlab}
\bibliographystyle{CSE}
